---
title: "STA 779: Final Project"
format: gfm
author: "Jackson Gazin, Ashley Mullan, Anh Nguyen"
date: '2022-03-29'
---

# Library

```{r}
#| label: load-pkgs
#| message: false
#| warning: false
# tidyverse
library(tidyverse)
library(gridExtra)
library(grid)
# dag
library(ggdag)
# eda
library(skimr)
library(visdat)
# plot
library(halfmoon)
# table 
library(gtsummary)
library(labelled)
library(broom)
library(survey)
# propensity
library(propensity)
# adjusting SEs
library(rsample)
library(sandwich)
# sensitivity
library(tipr)

```

```{r}
#setwd("~/final-project")
```

# Problem

Problem statement: Controlling blood pressure (BP) reduces the risk for cardiovascular disease. However, the prevalence of BP control (i.e., systolic BP < 140 and diastolic BP < 90) among US adults with hypertension has decreased since 2013. We invite teams to analyze publicly available data from US adults to help identify potential causes or correlates of worsening BP control among US adults with hypertension over the past decade, as this may allow for development of effective interventions to help control BP and prevent cardiovascular disease.

Description: Data from the US National Health and Nutrition Examination Survey (NHANES) containing information about demographics, blood pressure levels, hypertension status, antihypertensive medication usage, and co-morbidities for a sample of 59,799 noninstitutionalized US adults collected between 1999 and 2020.

# Data

```{r}
#devtools::install_github("jhs-hwg/cardioStatsUSA")
library(cardioStatsUSA)
data("nhanes_data")
```

```{r}
data(nhanes_key)
```


# EDA

```{r}
# shiny to explore data
app_run()
```


# Causal question

# Causal Diagram

from literature, ways to control blood pressure:

bp control ~ exercise + white coat effect + pharmacotherapy + diet + sodium + alcohol + smoking + sleep + stress + regular health check up

causal questions:

For noninstitutionalized US adults in the NHANES dataset collected between 1999 and 2020, does using cholesterol-lowering medication affects blood pressure control?

using variables from data set:

outcome: bp_sys_mean, bp_dia_mean (might only choose one)
exposure: chol_med_use

let demographics = age + race + gender

bp_sys_mean ~ bp_med_use + cc_smoke + cc_bmi + chol_total + demographics + chol_med_use 
bp_med_use ~ bp_sys_mean
cc_smoke ~ demographics 
cc_bmi ~ demographics 
chol_total ~ cc_bmi + demographics 
chol_med_use ~ chol_total 


```{r}
dag <- dagify(
  bp_sys_mean ~ bp_med_use + cc_smoke + cc_bmi + chol_total + demographics + chol_med_use, 
  cc_smoke ~ demographics,
  cc_bmi ~ demographics,
  chol_total ~ cc_bmi + demographics,
  chol_med_use ~ chol_total + demographics + cc_bmi + bp_med_use,
  exposure = "chol_med_use",
  outcome = "bp_sys_mean",
  labels = c(
    bp_sys_mean = "systolic blood pressure",
    bp_med_use = "antihypertensive medication",
    cc_smoke = "smoking",
    cc_bmi = "bmi",
    chol_total = "total cholesterol",
    demographics = "demographics",
    chol_med_use = "cholesterol-lowering medication"
  )
)


ggdag(dag, use_labels = "label", text = FALSE) +
  theme_dag()

ggsave("dag.png")
```


```{r}
ggdag_adjustment_set(
  dag,
  use_labels = "label",
  text = FALSE
) + theme_dag()
```


# Propensity score
```{r}
#check the vibes to think about an estimator

nhanes_data |> 
  drop_na() |>
  ggplot(aes(fill = chol_med_use, x = bp_sys_mean)) +
  geom_mirror_histogram()

data <- nhanes_data |> 
  as.data.frame() |> 
  drop_na() |>
  mutate(log_age = log(demo_age_years)) |>
  select(chol_med_use, chol_total, bp_sys_mean, bp_med_use, cc_bmi, log_age, demo_race, demo_gender)

prop <- glm(chol_med_use ~ chol_total + bp_med_use + cc_bmi + log_age + demo_race + demo_gender, 
    data = data,
    family = binomial())
data <- prop |> 
  augment(type.predict = "response", data = data)  |>
  select(-c(.resid, .hat, .sigma, .cooksd, .std.resid)) |>
  rename("prop" = ".fitted")
data <- data |>
  mutate(w_att = wt_att(prop, chol_med_use)) 

```

Check the Propensity Model
```{r}
uw_mirror <- data |> 
  ggplot(aes(x = prop, 
             group = chol_med_use, 
             fill = chol_med_use)) +
  geom_mirror_histogram(bins = 20) +
  theme_bw() +
  labs(x = "Propensity Score",
       y = "Count",
       fill = "Cholesterol Meds?",
       title = "No Weighting") +
  scale_fill_manual(values = c("#006241", "#cba258"))

w_mirror <- data |>
  ggplot(aes(x = prop, fill = chol_med_use)) +
  geom_mirror_histogram(
    aes(fill = chol_med_use,
        weight = w_att),
    binwidth = 0.05) +
  scale_fill_manual(values = c("#006241", "#cba258")) +
  scale_y_continuous(labels = abs) +
  theme_bw() +
  labs(x = "Propensity Score",
       fill = "Cholesterol Meds?",
       y = "Count",
       title = "ATT Weighting")

mirrors <- grid.arrange(uw_mirror, w_mirror, nrow = 2, top = textGrob("Mirrored Histograms"))
ggsave("mirrors.png", mirrors)

smds <- tidy_smd(
  data,
  .vars = c(chol_total, log_age, bp_med_use),
  .group = chol_med_use,
  .wts = w_att
)

love <- smds |>
  ggplot(aes(x = abs(smd), y = variable,
             group = method, color = method)) +
  geom_love() + 
  theme_bw() +
  labs(title = "Love Plot: Examining SMDs",
       x = "Absolute Value of Standard Mean Difference")

love #no problems here, weighting corrects SMD
ggsave("love.png", love)

w_ecdf_ct <- data |>
  ggplot(aes(x = chol_total, color = chol_med_use)) +
  geom_ecdf(aes(weights = w_att)) +
  scale_color_manual(
    "Uses Cholesterol Medication?",
    values = c("#FF69B4", "#013220"),
    labels = c("Y", "N")) +
    labs(x = "Total Cholesterol", y = "Prop <= X",
           title = "Weighted ECDF") +
  theme_bw() 

uw_ecdf_ct <- data |>
  ggplot(aes(x = chol_total, color = chol_med_use)) +
  geom_ecdf() +
  scale_color_manual(
    "Uses Cholesterol Medication?",
    values = c("#FF69B4", "#013220"),
    labels = c("Y", "N")) +
    labs(x = "Total Cholesterol", y = "Prop <= X",
           title = "Unweighted ECDF") +
  theme_bw() 

w_ecdf_age <- data |>
  ggplot(aes(x = log_age, color = chol_med_use)) +
  geom_ecdf(aes(weights = w_att)) +
  scale_color_manual(
    "Uses Cholesterol Medication?",
    values = c("#FF69B4", "#013220"),
    labels = c("Y", "N")) +
    labs(x = "Age", y = "Prop <= X",
           title = "Weighted ECDF") +
  theme_bw() 

uw_ecdf_age <- data |>
  ggplot(aes(x = log_age, color = chol_med_use)) +
  geom_ecdf() +
  scale_color_manual(
    "Uses Cholesterol Medication?",
    values = c("#FF69B4", "#013220"),
    labels = c("Y", "N")) +
    labs(x = "Log Age", y = "Prop <= X",
           title = "Unweighted ECDF") +
  theme_bw() 

ecdfs_choltot <- grid.arrange(uw_ecdf_ct, w_ecdf_ct, nrow = 2)
ecdfs_age <- grid.arrange(uw_ecdf_age, w_ecdf_age, nrow = 2)

ecdfs_choltot #only minor issues here
ecdfs_age
ggsave("ecdfs_choltot.png", ecdfs_choltot)
ggsave("ecdfs_age.png", ecdfs_age)

```

Outcome Model
```{r}
outcome_mod <- lm(bp_sys_mean ~ chol_med_use, data = data, weights = w_att)
robust_var <- sandwich(outcome_mod)[2, 2]
point_est <- coef(outcome_mod)[2]
lb <- round(point_est - 1.96 * sqrt(robust_var),3)
ub <- round(point_est + 1.96 * sqrt(robust_var),3)
paste0("The point estimate for the ATT is ", round(point_est,3), " with 95% CI (", lb, ",", ub,").")
```

# Sensitivity analysis
```{r}
tip_coef_with_continuous(
  effect_observed = -2.71, #upper bound of CI
  exposure_confounder_effect = c(0.05, 0.10, 0.15),
  verbose = FALSE
)
```
Verbiage for sensitivity:
The estimated effect assumes a correct DAG. However, it is possible that we did not include all possible confounding variables in the DAG. To counter this, we run sensitivity analysis on the estimated ATT. Specifically, we compute the necessary relationship between an unmeasured confounder and the outcome of weight change required to tip the lower bound of the confidence interval to the null at the 5% level. We try exposure confounder effects of size 0.05, 0.10, and 0.15, and these require confounder-outcome effects of BLAH to tip the interval. As the smallest of these is over 6 times the estimated effect, it is reasonable to say that the ATT we computed is resistant to confounding.


#Causal Question Verbiage
Among those who take cholesterol medication, does stopping the medication have an effect on systolic blood pressure?


