---
title: Average treatment effect of cholesterol-lowering medication and average systolic blood pressure (SBP), mm Hg.

author: 
- name: Jackson Gazin
  affiliation: Department of Statistics, Wake Forest University
  email: gazij22@wfu.edu
  
- name: Ashley Mullan
  affiliation: Department of Statistics, Wake Forest University
  email: mullae22@wfu.edu
  
- name: Anh Nguyen
  affiliation: Department of Statistics, Wake Forest University
  email: nguyp22@wfu.edu

keywords:
- cholesterol-lowering medication
- systolic blood pressure
- blood pressure control

abstract: |
  We investigate the average treatment effect among the treated (ATT) of cholesterol-lowering medication on the mean systolic blood pressure (mm Hg). Using data from the National Health and Nutrition Examination Survey (NHANES), we fit a propensity score model to estimate the ATT among adults living in the United States of America. 

bibliography: bibliography.bib
output: rticles::biometrics_article
month: "`r format(Sys.Date(), '%b')`"
year: "`r format(Sys.Date(), '%Y')`"
referee: true # Papers submitted to Biometrics should ALWAYS be prepared
              # using the referee option!!!! Turn off only to preview
              # two column-format
---

# Introduction {#intro}

Controlling blood pressure (BP) reduces the risk for cardiovascular disease. However, the prevalence of BP control (i.e., systolic BP < 140 and diastolic BP < 90) among US adults with hypertension has decreased since 2013. We invite teams to analyze publicly available data from US adults to help identify potential causes or correlates of worsening BP control among US adults with hypertension over the past decade, as this may allow for development of effective interventions to help control BP and prevent cardiovascular disease.

# Materials and methods {#methods}

## Data

The National Health and Nutrition Examination Survey (NHANES) combines interviews and physical examinations to assess the health and nutritional status of adults and children in the United States of America. The program started in the early 1960s and has been conducted every two years since 1999. The survey samples from a nationally representative 5,000 persons each year. The participants are located in counties across the country, 15 of which are visited each year. The interview asks questions about demographic, socioeconomic, dietary, and health-related questions. The examination consists of medical, dental, physiological measurements, and laboratory tests.

The NHANES dataset we are using can be downloaded from __cite__ . The dataset contains information from the survey from 1999 to 2020 with a sample of 59,799 rows and 111 chosen columns. Each row is a noninstitutionalized US adults participated in the survey between 1999 and 2020. The columns contain information about demographics, blood pressure levels, hypertension status, antihypertensive medication usage, and co-morbidities. 

For this analysis, __missing data__

## Statistical methods

We fit a propensity score model on 

### Exploratory data analysis

### Modeling

# Results {$results}

## Study population

## Propensity score model

#### Diagnostics

## Average treatment effect among the treated

# Discussion {#discussion}

# Supplementary information {#supplement}

The data can be downloaded from GitHub or accessed via the cardioStatsUSA R package. For both the file and information about the R package, see https://github.com/jhs-hwg/cardioStatsUSA.

All code for the analysis can be accessed at __link github__

# Acknowledgement {#acknowledge}

# Section title {#sec:1}

Text with citations by @heagerty2000time, [@pepe2003statistical].

## Subsection title {#sec:2}

as required [@hoerl1970ridge; @zou2005regularization]. Don't forget to give each section
and subsection a unique label (see Sect. \ref{sec:1}).

#### Paragraph headings 

Use paragraph headings as needed.

## Equations

Here is an equation:

$$ f_{X}(x) = \left(\frac{\alpha}{\beta}\right)\left(\frac{x}{\beta}\right)^{\alpha-1}e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; \alpha,\beta,x > 0 $$

Here is another:
\begin{align}
a^2+b^2=c^2
\end{align}

Inline equations: $\sum_{i = 2}^\infty\{\alpha_i^\beta\}$

# Figures and tables


## Figures coming from R

#### Normal figure embedded in text

```{r fig2, fig.width = 4, fig.height = 4, fig.cap = "Output from `pdf()`", echo = FALSE, dev = "pdf"}
plot(runif(25) ~ runif(25))
```

\clearpage

## Tables coming from R


```{r tab1, results = 'asis'}
print(xtable::xtable(head(mtcars)[,1:4], 
caption = "Caption centered under table", label = "tab1"), 
comment = FALSE, timestamp = FALSE, caption.placement = "top")
```

Table \ref{tab1} shows these numbers. Some of those numbers are plotted in Figure \ref{fig:fig1}.


```{r tab2}
head(mtcars[,1:4])
```


# References
